\input{../latex/shared}

\renewcommand{\course}{Maths for Intelligent Systems}
\renewcommand{\coursedate}{Summer 2019}

\renewcommand{\exnum}{Exercise 1}

\exercises

\excludecomment{solution}

\exercisestitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% \exsection{Examples for Vector Spaces (voluntary, discussed on Oct 23rd)}

%% \begin{enumerate}
%%   \item Find 3 interesting examples of vector spaces that we haven't covered in
%%     class.  What are their dimensions? What bases are commonly used to
%%     represent them?  (It's okay to look this up in a textbook or online---just
%%     find interesting examples.)

%%   \item Come up with one example of a set which is \emph{almost} a vector
%%     space, i.e.\ it satisfies some of the core requirements of a vector space,
%%     but not all.

%%   \item Consider the set of of all \emph{even} functions.  Is this a Vector
%%     space?  If so, define 3 different bases which span the
%%     entire space.

%% \end{enumerate}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{Basis}

Given a linear transform $f: \RRR^2 \to \RRR^2,$
$$ f(x) = A x = \mat{cc}{7 & -10 \\ 5 & -8} x ~.$$ Consider the basis
$\BB = \big\{ \mat{c}{1\\1}, \mat{c}{2\\1} \big\}$, which we also
simply refer to by the matrix $B=\mat{cc}{1 & 2 \\ 1 & 1}$. Given a
vector $x$ in the vector space $\RRR^2$, we denote its coordinates in
basis $\BB$ with $x^B$.

\begin{enumerate}
\item Show that $x = B x^B$.


\item What is the matrix $F^B$ of $f$ in the basis $\BB$, i.e., such that $[f(x)]^B = F^B x^B$? Prove the general equation $F^B = B^\1 A B$.


\item Provide $F^B$ numerically
\end{enumerate}

Note that for a matrix $M = \mat{cc}{a & b \\ c & d},~ M^\1 = \frac{1}{ad-bc}\mat{cc}{d & -b \\ -c & a}$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{From the Robotics Course}

You have a book lying on the table. The edges of the book define the basis
$B$, the edges of the table define basis $A$. Initially $A$ and $B$ are identical (also their origins align). Now we rotate the book
 by $45^\circ$ counter-clock-wise
about its origin.

\begin{enumerate}
\item Given a dot $p$ marked on the book at position
$p^B=(1,1)$ in the book coordinate frame, what are the coordinates
$p^A$ of that dot with respect to the table frame?

\item Given a point $x$ with coordinates $x^A=(0,1)$ in table
frame, what are its coordinates $x^B$ in the book frame?


\item What is the \emph{coordinate} transformation matrix from book frame to table frame, and from table frame to book
frame?

\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{Bases for Polynomials}

Consider the set $V$ of all polynomials $\sum_{i=0}^n \alpha_i x^i$ of
degree $n$, where $x \in \RRR$ and $\alpha_i \in\RRR$ for each
$i=0,\ldots,n$.

\begin{enumerate}
\item Is this set of functions a vector space? Why?



\item Consider two different bases of this vector space: 
\begin{align*}
  \mathcal{A} = \{1,\ x,\ x^2,\ \ldots,\ x^n\}
\end{align*}
and 
\begin{align*}
  \mathcal{B} = \{1,\ 1 + x,\ 1 + x + x^2,\ \ldots,\ 1 + x + \ldots + x^n\}.
\end{align*}
Let $f(x) = 1 + x + x^2 + x^3$ be one element in $V$. (This function $f$ is a vector in the
  vector space $V$, so from here on we refer to it as a vector rather
than a function.)

What are the coordinates $[f]^A$ of this vector in basis $\mathcal{A}$?

What are the coordinates $[f]^B$ of this vector in basis $\mathcal{B}$?


\item What matrix $I^{BA}$ allows you to convert between coordinates $[f]^A$ and
  $[f]^B$, i.e.\ $[f]^B = I^{BA} [f]^A$?  Which matrix $I^{AB}$ does the same
  in the opposite direction, i.e.\ $[f]^A = I^{AB} [f]^B$?  What is the
  relationship between $I^{AB}$ and $I^{BA}$?


\item What does the difference between coordinates $[f]^A - [f]^B$
        represent?


\item Consider the linear transform $t:~ V \to V$ that maps basis elements of $\mathcal{A}$ directly to basis elements of 
$\mathcal{B}$:
\begin{align*}
  &1 \rightarrow 1 \\
  &x \rightarrow 1 + x \\
  &x^2 \rightarrow 1 + x + x^2 \\
  &\vdots \\
  &x^n \rightarrow 1 + x + x^2 + \cdots + x^n 
\end{align*}

\begin{itemize}
\item What is the matrix $T^A$ for the linear transform $t$ in the basis $\AA$, i.e., such that $[t(f)]^A = T^A [f]^A$? (Basis $\AA$ is used for both, input and output spaces.)
\item What is the matrix $T^B$ for the linear transform $t$ in the basis $\BB$, i.e., such that $[t(f)]^B = T^B [f]^B$? (Basis $\BB$ is used for both, input and output spaces.)
\item What is the matrix $T^{BA}$ if we use $\AA$ as input space basis, and $\BB$ as output space basis, i.e., such that $[t(f)]^B = T^{BA} [f]^A$?
\item What is the matrix $T^{AB}$ if we use $\BB$ as input space basis, and $\AA$ as output space basis, i.e., such that $[t(f)]^A = T^{AB} [f]^B$?
\item Show that $T^B = I^{BA} T^A I^{AB}$ (cp.\ Exercise 1(b)). Also note that $T^{AB} = T^A I^{AB}$ and $T^{BA} = I^{BA} T^A$.
\end{itemize}


%% \item The identity transformation $I$ maps $f\mapsto f$.  What
%%    is the matrix $M$ which performs the identity transformation in all possible combinations of input and output basis? i.e. find the matrices $M$ such that
%%        \begin{itemize}
%%           \item $[If]^A = M [f]^A$
%%           \item $[If]^B = M [f]^B$
%%           \item $[If]^B = M [f]^A$
%%           \item $[If]^A = M [f]^B$
%%         \end{itemize}


\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\exerfoot
