\input{../latex/shared}

\renewcommand{\course}{Robotics}
\renewcommand{\coursepicture}{roboticsLecture}
\renewcommand{\coursedate}{Winter 2014}
\renewcommand{\exnum}{2}

\exercises

\excludecomment{solution}

\exercisestitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{Geometry}

a) Read the notes on basic 3D geometry at
{\tiny\url{http://ipvs.informatik.uni-stuttgart.de/mlr/marc/notes/3d-geometry.pdf}} at
least until section 2. Prepare questions for the exercises if you have any.

b) You have a book (coordinate frame $B$) lying on the table (world
frame $W$). Initially $B$ and $W$ are identical. Now you move the book 1
unit to the right, then rotate it by $45^\circ$ counter-clock-wise
around its origin. Given a dot $p$ marked on the book at position
$p^B=(1,1)$ in the book coordinate frame, what are the coordinates
$p^W$ of that dot with respect to the world frame?

c) Given a point $x$ with coordinates $x^W=(0,1)$ in world
frame, what are its coordinates $x^B$ in the book frame?

d) What is the \emph{coordinate} transformation from world frame to book
frame, and from book frame to world frame?

Please use homogeneous coordinates to derive these answers.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{Subscribe to the Email list}

Please subscribe here:\\
{\small\url{https://mailman.informatik.uni-stuttgart.de/mailman/listinfo/lehre14-rob}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{Simulation software}

Future exercises will require to code some examples in C/C++. Test if you
can compile and run the lib that accompanies this lecture. Report
on problems with installation.

On Ubuntu:
\begin{itemize}
\item install the packages \\ {\small @liblapack-dev@ @freeglut3-dev@
@libqhull-dev@ @libf2c2-dev@ @libann-dev@ @gnuplot@ @doxygen@ @libglew1.6-dev@}
\item get the code
from \\ {\small\url{http://ipvs.informatik.uni-stuttgart.de/mlr/marc/source-code/libRoboticsCourse.14.tgz}}
\item {\small
@tar xvzf libRoboticsCourse.14.tgz@\\
@cd share/examples/Ors/ors@\\
@make@\\
@./x.exe@

}
\end{itemize}




\exerfoot
