\input{../latex/shared}

\renewcommand{\course}{Robotics}
\renewcommand{\coursepicture}{roboticsLecture}
\renewcommand{\coursedate}{Winter 2014}
\renewcommand{\exnum}{3}

\exercises

\excludecomment{solution}

\exercisestitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{Task spaces and Jacobians}

%% a) Assume you want to control the position of the left hand ($L$)
%% relative to the right hand ($R$), that is $p^R$, where $p$ is a point
%% on the left hand and $p^R$ its coordinate w.r.t.\ the right hand's
%% frame. What is the respective task map $\phi:~ q \mapsto p^R$
%% expressed on the basic transformations $\TR{W}{L}$ and $\TR{W}{R}$? What is
%% its Jacobian?

In the lecture we introduced the basic kinematic maps
$\phi^\pos_{i,v}(q)$ and $\phi^\veC_{i,v}(q)$, and their Jacobians,
$J^\pos_{i,v}(q)$ and $J^\veC_{i,v}(q)$, where $i$ may refer to any
part of the robot and $v$ is any point or direction on this part. In
the following you may assume that we have routines to compute these
for any $q$. The problem is to express other kinematic maps and their
Jacobians in terms of these knowns. In the following you're asked to
define more involved kinematics maps (a.k.a.\ task maps) $\phi(q)$ to solve
certain motion problems. Please formulate all these maps such that the
overall optimization problem is
$$ f(q) = \norm{q-q_0}^2_W + \norm{\phi(q)}^2 $$
that is, the motion aims to minimize $\phi(q)$ to zero (absorb the
$y^*$ in the definition of $phi$.)

a) Assume you would like to control the pointing direction of the robot's
head (e.g., its eyes) to point to a desired external world point $x^W$. What
task map can you define to achieve this? What is the Jacobian?


b) You would like the two hands or the robot to become parallel
(e.g.\ for clapping). What task map can you define to achieve this?
What is the Jacobian?


c) You would like to control a standard endeffector position $p_\eff$
to be at $y^*$, as usual. Can you define a 1-dimensional task map
$\phi:~ \RRR^n \to \RRR$ to achieve this? What is its Jacobian?


%% \exsection{Derivation of inverse kinematics}

%% \show[.25]{optim-InvKin}

%% We want to find an optimum for joint angle step $\d q\in\RRR^n$
%% under two ``competing tasks'': On the one hand, we want $\d q$ to be
%% small (be lazy); on the other hand, we want $\d q$ to be close to
%% the hyperplane $J\d q-\d y=0$, for given $J\in\RRR^{d\times n}$
%% and $\d y\in\RRR^d$. We associate square costs to both tasks and
%% define the cost function
%% \begin{align} \label{eq1}
%% f(\d q)
%%  &= \d q^\T~ W~ \d q + (J\d q-\d y)^\T~ C~ (J\d q-\d y) ~.
%% \end{align}
%% We introduced metrics (=symmetric positive definite matrix)
%% $W \in\RRR^{n\times n}$ and $C \in \RRR^{d\times d}$ to weight the
%% cost terms. Find the optimal $\d q^* = \argmin_{\d q} f(\d q)$.

%% \emph{Simplification:} If you like, first solve a simplified problem: 
%% Drop $W$ and $C$ (let $W$ and $C$ be Euclidean metrics).

%% \emph{Bonus:} What is the solution for $C \to \infty\Id$ (enforcing
%% the solution to be definitely on the hyperplane)? Tip: Before taking the limit,
%% rewrite the solution using the \emph{Woodbury identity} (for positive
%% definite matrices $W$ and $C$):
%% \begin{align}
%% (W + J^\T C J)^\1 J^\T C = W^\1 J^\T (C^\1 + J W^\1 J^\T)^\1 ~.
%% \end{align}

%% \emph{For discussion during exercises:} How are the solutions related to the
%% so-called pseudo-inverse? How are they related to singularities, the
%% ``singularity robust Jacobian'' and ``Tikhonov regularization''?

\exsection{IK in the simulator}

Download the simulator code from
{\tiny\url{http://ipvs.informatik.uni-stuttgart.de/mlr/marc/source-code/libRoboticsCourse.14.tgz}}. (See
last exercise for instructions. The @libRoboticsCourse.13.msvc.tgz@
might include a project file for MSVC.)  The header
@<src/Ors/roboticsCourse.h>@ provides a very simple interface to the
simulator---we will use only this header and some generic matrix
functionalities. In @share/examples/Core/array@ you find many examples
on how to use the @arr@ for working with vectors and matrices (many
conventions are similar to Matlab).

Consider the example in @teaching/RoboticsCourse/01-kinematics@
(rename @main.problem.cpp@ to @main.cpp@). The goal is to reach the
coordinates $y^*=(-0.2, -0.4, 1.1)$ with the right hand of the
robot. Assume $W=\Id$ and $\s=.01$.

a) The example solution generates a motion in one step using inverse
kinematics $\d q = J^\sharp~ \d y$ with $J^\sharp = (J^\T J + \s^2
W)^\1 J^\T$. Record the task error, that is, the deviation of hand
position from $y^*$ after each step. Why is it initially large?

b) Try to do 100 smaller steps $\d q = \a J^\sharp~ \d y$ with $\a=.1$
(each step starting with the outcome of the previous step). How does
the task error evolve over time?

c) Generate a nice trajectory composed of $T=100$ time
steps. Interpolate the target linearly $\hat y \gets y_0 + (t/T)(y^*-y_0)$ in
each time step.

d) Generate a trajectory that moves the right hand in a circle centered at
$(-0.2, -0.4, 1.1)$, aligned with the $xz$-plane, with radius $0.2$.


\exerfoot
