\input{../latex/shared}

\renewcommand{\course}{Robot Learning}
\renewcommand{\coursedate}{Summer 2024}
\renewcommand{\exnum}{Weekly Exercise 11}

\renewcommand{\teacher}{Marc Toussaint \& Wolfgang H{\"o}nig}
\renewcommand{\addressTUB}{
  Learning~\&~Intelligent~Systems Lab, Intelligent Multi-Robot Coordination Lab, TU~Berlin\\\small
  Marchstr. 23, 10587 Berlin, Germany
}

\exercises

\ifthenelse{\isundefined{\scripthead}}{
  \usepackage{tikz}
  \usetikzlibrary{shapes,snakes}
  \usepackage{tkz-base}
  \usepackage{tkz-euclide}
}{}

\input{macros-local}

\excludecomment{solution}

\exercisestitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{Literature: Neural-Swarm2}

Here is the paper we discussed in the lecture that uses (and extends) deep sets for a control problem that arises in multi-robot aerial swarms~\footnote{A shorter and perhaps easier to follow earlier work is
\bibentry{2020-shi-NeuralSwarmDecentralizedCloseProximity}
}:

\bibentry{2022-shi-NeuralSwarm2PlanningControl}

The paper is an extension of the NeuralLander paper to the multi-robot case we discussed in exercise 8. 

Questions:
\begin{enumerate}
\item How does the dataset exactly look like? How was the data obtained? What sensing/measurement capabilities were needed to obtain such data?


\item Write down pseudo code on how one can use SGD or Adam and train a 2-group permutation-invariant function using the heterogeneous deep sets proposed in (9).


\item Consider the use-case of motion planning (Fig. 6). Explain how the neural network is applied inside the motion planner.


\item In the considered examples for $K$-group permutation invariant functions, $K$ is relatively small (4 in the paper). Consider the case where $K$ is large or unknown, for example if we are able to measure the size of the neighboring robot (a real value). How could learning be applied in this case?


\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exsection{Encodings for Environmental Monitoring}

Consider a team of robots that is spatially distributed as shown below.
In the figure, circles are robots, the numbers are their associated measurements (such as temperatures), and lines indicate the existence of a communication link.
The goal is to find the minimum of their sensor measurements.
In this question we will explore various concrete encodings for such problem.

\begin{center}
  \begin{tikzpicture}
      \tkzInit[xmax=5,ymax=5,xmin=0,ymin=0]
      %\tkzGrid
      \tkzAxeXY

      \node[circle,
          fill=black,
          label=$5$] (n1) at (1, 1) {};

      \node[circle,
          fill=black,
          label=$2$] (n2) at (4, 5) {};

      \node[circle,
          fill=black,
          label=$\,\,\,\,\,3$] (n3) at (4, 3) {};

      \node[circle,
          fill=black,
          label=$4$] (n4) at (2, 3) {};

      \draw (n1) -- (n4);
      \draw (n4) -- (n3);
      \draw (n4) -- (n2);
      \draw (n2) -- (n3);
  \end{tikzpicture}
  \end{center}

\begin{enumerate}
\item First consider the abstract, centralized setting with function $f(\mathcal X) = \min_{x\in\mathcal X} x$, where $\mathcal X$ is a set of real numbers. In other words, the function takes one or more numbers as input and returns the smallest element of these numbers. Recall that the deep set 
\begin{equation}
f(\mathcal X) \approx \rho \left(\sum_{x\in\mathcal X} \phi(x) \right)
\end{equation}
should be able to approximate this function. Provide concrete (differentiable) functions for $\rho$ and $\phi$ for this case.\\
Hint: You can find some inspiration in the original Deep Set paper or the paper from question 1.


\item Now assume the case where robots have a limited communication radius. One example is shown in the figure, where the lines show communication links. Define the \texttt{Aggregate} and \texttt{Update} functions of a simple message-passing neural network.

Demonstrate in the example above, how the node at $(1,1)$ computes the minimum value.


% min(5,4) -> 4
% (4,3,2) -> 2

%TODO: concrete instance; ask to manually show execution

% 

\item How could a CNN be used for the case with limited communication radius? Be specific about the layers the CNN should have.


\item For the use-case outlined above, what are advantages and disadvantages of the three encodings (Deep Sets, GNN, CNN)? Consider both small (=few neighbors) and large (=many neighbors) cases.


\end{enumerate}

% example problem -> playing with different potential network architectures (CNN, DeepSet, GNN)

% deepset on paper for something like the max() function?

% GNN on paper for the same function?

% CNN on paper for the same function

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \exsection{Programming}

% To be decided on Monday, if there will be another programming assignment.

% Either train a simple DeepSet or GNN using pytorch, in a basic robotic domain (flocking?, collision avoidance?, perhaps imitating artificial potentials? or plain task assignment [not very robotic])

% Variant 1:
% team of 3 double integrators -> expert using cvxpy (stacked) -> find distributed policy, e.g, with deepset

% Variant 2:
% 10 robots, 10 tasks, Hungarian method as expert
% GNN that learns how to assign tasks


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\ifthenelse{\isundefined{\scripthead}}{
\bibliographystyle{plainurl-lis}
\bibliography{b8-MultiRobotLearning}
}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\exerfoot
