\input{../latex/shared}
\note

\title{Lecture Note:\\ Robot Kinematics \& Dynamics}
\author{Marc Toussaint\\\small Learning \& Intelligent Systems Lab, TU Berlin}

\makeatletter
\renewcommand{\@seccntformat}[1]{}
\makeatother

\newcommand{\pa}{\text{par}}
\newcommand{\SO}{\text{SO}}
\newcommand{\SE}{\text{SE}}
\newcommand{\se}{\text{se}}
\newcommand{\ang}{\text{ang}}

\notetitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This is meant as essentials on robotic kinematics and
dynamics -- developed as background for the \emph{Robot Learning} course.

\subsection{Articulated Multibody System}

A robot is a multibody
system. Each body has a \textbf{pose} $x_i\in\SE(3)$, an inertia $(m_i, I_i)$
with mass $m_i\in\RRR$ and inertia tensor $I_i \in \RRR^{3\times 3}$
sym.pos.def., and a shape $s_i$ (any shape representation that defines
a pairwise signed-distance $d(s_i, s_j)$ is sufficient).

We assume this multibody system is tree-structured, i.e., every body
is linked to a parent body or the world. Body $i$ has a relative
transformations $Q_i \in \SE(3)$ from its parent (or world)
$\pa(i)$.  Given the tree
structure, we can compute the pose $x_i$ of each body simply by
forward chaining all relative transformations all $Q_j$ from world to
$i$. Some robotic systems might not really be tree-structured --
these will first be represented as a tree
and then additional constraints to describe loops are added.

What is special about robots is that some of the relative
transformations have degrees of freedom (dofs) that are
\emph{articulated} (i.e., are motorized/movable). Let $Q_i$ have 1
dofs $q_i \in \RRR^1$, then $Q_i(q_i)$ is a function of this dof. We
stack all dofs of the whole multibody tree into a vector $q\in\RRR^n$,
which is called \textbf{joint vector}. We will discuss in more depth
the concepts of generalized and minimal coordinates below.


\subsection{Forward Kinematics \& Jacobian}

We have defined $x_i\in\SE(3)$ as the pose of body $i$, and
$q\in\RRR^n$ as the dofs of the full multibody system. The mappings $\phi_i: q
\mapsto x_i$ (for any $i$) is what so-called ``forward kinematics'' is
concerned about. We already explained that $x_i = \phi_i(q)$ can be
computed simply by chaining all relative transformations. We call this
the (full) forward kinematics.\footnote{Very often we are not interested to
really compute all poses $x_i$ of the multibody system, but only the
pose of one relevant body $i$ (esp.\ the so-called endeffector or
manipulator), or only the position $x^\pos_i$ or some rotated vector $x^v_i
= R_i v$ (with $R_i \in \SO(3)$ its orientation) for body
$i$. In robotics, the word forward kinematics is used often to refer
only to compute the endeffector pose, position, or orientation.}

The derivative $J_i(q)=\del_q \phi_i(q) ~ \in \se(3) \otimes \RRR^n$ is
of central importance to later solve constraint problems, or also to
relate joint space velocities $\dot q$ to the velocity $\dot x_i \in
\se(3)$ of the $i$th body. Note that elements $(v,w)\in \se(3) \equiv
\RRR^6$ are 6-vectors composed of the linear velocity $v\in\RRR^3$ and
angular velocity $w\in\RRR^3$ (always in world
coordinates). Therefore, we can write $J_i(q) \in \RRR^{6 \times n}$
as a matrix, called \textbf{Jacobian}, and

$$(v_i,w_i) = J_i(q) \dot q$$

gives us the linear and angular velocity of body $i$ when joints have
velocities $\dot q$. In practice, code typically returns separate positional Jacobian $J_i^\pos \in \RRR^{3\times n}$ and angular
Jacobian $J_i^\ang \in \RRR^{3\times n}$. In fact, the core job of a
kinematics engine is (1) to represent the articulated multibody tree,
(2) to forward compute $\phi_i(q)$, and (3) to compute
$J_i^\pos, J_i^\ang$ for any $i$.

Since we know how to compute $\phi_i(q)$, we could use ``autodiff'' to
also compute the derivative $J_i(q)$. However, the columns of the positional and
angular Jacobians can actually be computed very easily and more efficiently by simple
insight of how the local translational/angular velocity of each
joint dof translates to the translational/angular velocity of body
$i$.\footnote{For instance, if $j$ is a rotational (``hinge'') joint
around axis $e$ in the joint's origin frame $x_{\pa(j)}$, and body
$i$ is downstream at current pose $x_i$, then the $j$th column of
$$J^\ang_i$$ is $a = (R_{\pa(j)} e)$ (rotates about the axis of
$j$ in world coordinates), and the $j$th column of $$J^\pos_i$$ is
$$a \times (x^\pos_i - x^\pos_{\pa(j)})$$ (translates with a lever around
the axis). Similar arguments can be made if $j$ is a translational
(``prismatic'') joint, and a bit more complicated arguments if $j$ is
a ball joint parameterized by a quaternion $q_j \in\RRR^4$. Thinking
of other joint types as compositions of these basic joints, this
covers all cases.
}
The Jacobians are typically sparse for large robotic systems (e.g.,
multi-robot systems): Every column of $J_i\in\RRR^{6\times n}$
describes how dof $j\in\{1,..,n\}$ influences body $i$. This column
will be zero if dof $j$ is not between the world and body $i$ in the
tree.

\subsection{Fundamental Kinematics Concepts}

The word ``kinematics'' in general refers to the mathematical
description of the possible motions of a (potentially constrained) multibody
system or mechanism \emph{without considering the forces}.

For a multibody system, the poses $x_{1:m}$ fully describes the
\emph{configuration} of the system. When $x$ is some (potentially
redundant) description of system state, we generally call its
embedding space the \textbf{configuration space} $\XX$. E.g., for our
multibody system $\XX=\SE(3)^m$ is a generic embedding space. However, not all
configurations are possible, e.g.\ because bodies are linked in a multibody
system, body shapes cannot penetrate, or obstacles block parts of the
configuration space. We can imagine the \textbf{feasible config.\ space}
$\XX_\text{fea}$ as a manifold of feasible configurations, potentially
with disconnected components or holes. (Actually also
trans-dimensionality, where some parts have different dimensionality is
possible, but we neglect this here.) When introducing 
coordinates $q\in\RRR^n$ for the feasible space these are called
\textbf{generalized coordinates}. (This should not be confused with the word
\textbf{canonical coordinates}, which is used for coordinates in the phase
space of a system, and usually denoted $(q,p)$.)

We discussed the manifold of feasible
configurations, however kinematics is about describing feasible
\emph{motions} on that manifold. Therefore, formally, kinematics
describes which $\dot q \in T_q \XX$ (in the tangent space of $\XX$)
are feasible, and therefore which paths of motion on the manifold.

A holonomic constraint is of the form $h(q, t)=0$, with $h$ a $d$-dimensional
function (where $t$ allows a dependence
on absolute time, which is hardly ever relevant in our field). In such a system,
the generalized coordinates were not minimal and provide only an
embedding space for the true, lower-dimensional feasible manifold. The
true \textbf{degrees of freedom} of the system are $p=n-d$. \textbf{Minimal
  coordinates} are defined to be generalized coordinates of dimension
$n=p$ (where no further holonomic constraints
exist).\footnote{Sometimes it may be convenient to stick with
non-minimal generalized coordinates: When the feasible manifold is
$S^1$, it might be convenient to use redundant $q=(x,y)$ coordinates
and add the constraint $x^2+y^2=1$ rather than
introducing an angle coordinate and running into the annoying modulo issue. Analogous
for $\SO(3)$ and embedding quaternion coordinates $\RRR^4$ may be more
convenient than minimal coordinates for $\SO(3)$. Further, when we
have a closed loop robotic system, it is convenient to use non-minimal
joint coordinates along a tree and profit from the efficiency of tree-based
kinematics engines, and handle the closed loop constraint
otherwise.}

A non-holonomic constraint is of the form $h(q,\dot q,t)=0$, which is
a general description of any constraints on
possible motions on the feasible manifold. A typical example is a car
or wheel in 2D: We describe the configuration naturally with $x =
(p,\varphi)$ with 2D position $p\in\RRR^2$ and heading angle
$\varphi\in\RRR$. Note that (without obstacles) any configuration is
feasible, so the full configuration space is feasible. As there is no
better alternative, we choose generalized coordinates equally as $q =
x$. However, at any $q$, the positional velocity is constraint to
aligned with the heading direction, $\dot p^\T (\sin\p, \cos\p) = 0$,
which is a non-holonomic constraint.

There are cases where a constraint is naturally expressed as $h(q,\dot
q,t)=0$ and ``looks'' non-holonomic, but actually it is so-called
integratable. This means that, by analyzing integrals of trajectories of
feasible velocities $\dot q(t)$, we understand that the actually
reachable $q$ all lie on a sub-manifold which we could more directly
be described by a holonomic constraint $h(q, t)=0$ (here, the absolute time
$t$ dependence really is important). So, such ``integratable
non-holonomic constraints'' can be reformulated to become holonomic
still describe a holonomic system. The literature describes elaborate
maths (Pfaffian form of constraints) to uniquely decide whether a system is truly
holonomic or non-holonomic. But this is rarely relevant for typical
robotic systems in our field.

\subsection{``Force Kinematics''}

We learned that with 
$(v_i,w_i) = J_i \dot q$ the Jacobian relates joint velocities to body
velocities. Let's do the analogous for forces: Assume we have a wrench
$(f_i,\tau_i)$ directly acting on body $i$, what ``do we feel'' in the
joints, i.e., what torques $u\in\RRR^n$ are propagated into the
joints? The answer is the Jacobian transpose: $u = J^\T
(f_i,\tau_i)$.\footnote{We can derive this by conserved work, or better,
power (=work/time): For joint torques $u$ and velocities $\dot q$ we
would consume power $u^\T \dot q$, which needs to equal the power
received at the body, $(f_i,\tau_i)^\T (v_i,w_i) = (f_i,\tau_i)^\T J
\dot q$. As this holds for any $\dot q$ we have $u^\T = (f_i,\tau_i)^\T J$.}


\subsection{Dynamics}

While kinematics describes which $q$ and $\dot
q$ are feasible, dynamics describes which $\ddot q$ are feasible. For
a passive dynamical system there is just one $\ddot q$ feasible: the
one that follows from the laws of physics. For an articulated robotic
system we can choose to exert torques $u$ in each joint (or some joints), and
thereby create various $\ddot q$. For standard, fully actuated robotic systems
we can command torques $u\in\RRR^n$ in all dofs and create arbitrary
accelerations. However, when our multibody system description includes
passive objects of the environment, or describes a free-floating
(walking/running) robot, the accelerations that can be generated for
all dofs (objects, free-floating body) are highly constrained,
depending esp.\ on contacts and possibilities of force transmission
to objects or the ground through contacts.

Assume we know how we want to accelerate $\ddot q$ the system, and
want to compute the necessary joint torques $u$ to achieve this acceleration. That
is, we want to derive the mapping $(q,\dot
q, \ddot q) \mapsto u$. Given the
Jacobians described above, this is easy to derive:
For each body, we can compute
$(v_i,w_i) = J_i \dot q$ and $(\dot v_i, \dot w_i) = J_i \ddot
q$. I.e., we know how we want each body to accelerate. The
Newton-Euler equation tells us such an acceleration would raise the
following inertia forces at the body:

\begin{align*}
\mat{c}{f_i \\ \tau_i}
&= \mat{c}{m_i \dot v_i \\\bar I_i \dot w_i + w_i\times \bar I_i w_i}
 = M_i J_i \ddot q + c_i ~,
\end{align*}

with $M_i = \diag(m \Id_3, \bar I_i)$, $c_i = (0_3, w_i\times \bar I_i
w_i)$, where $\bar I_i = R_i I_i R_i^\T$ and $I_i$ the inertia tensor in
body coordinates.

Conversely, to counteract these inertia forces we have to apply joint torques
$u = J_i^\T \[M_i J_i \ddot q + c_i\]$ -- this is how ``we feel'' the
inertial in the joints. We can separately consider
gravity: By the same argument we need joint torques $u = J_i^\T g_i$
to counteract gravity, where $g_i = (-m g e_z, 0_3)\in\RRR^6$ has only a
single entry in $-z$ direction. As we have many bodies that are accelerated and need
gravity compensation, we overall have

$$u = \sum_{i=1}^m J_i^\T \[M_i J_i \ddot q + c_i + g_i\] ~.$$

Other texts provide much more lengthy derivations either via the general
Euler-Lagrange equations, or recursive Newton-Euler equations. I find
the above very concise and easy to implement, and efficient with
sparse $J_i$. The first term $\sum_{i=1}^m J_i^\T M_i J_i \ddot q$ can elegantly
also be found in the Euler-Lagrange derivation,%
\footnote{The Euler-Lagrange derivation starts with
$\frac{d}{dt} \frac{\del L}{\del \dot q} - \frac{\del L}{\del q} = u$,
where $L(q,\dot q) = T(q,\dot q) - U(q)$ with the system kinetic energy
$$T(q,\dot q) = \sum_i \half m_i v_i^2 + \half w_i^\T \bar I_i w_i
= \sum_i \half \dot q^\T J_i^\T M_i J_i \dot q,~
M_i = \diag(m_i\Id_3, \bar I_i), $$
and the system potential energy $U(q) = \sum_i g m_i x_i^\text{z}$. When computing the partial derivatives analytically we get something
of the form
$$
u = \frac{d}{dt} \frac{\del L}{\del \dot q} - \frac{\del L}{\del q} 
 = M(q) \ddot q + \dot M \dot q - \frac{\del T}{\del q} + \frac{\del U}{\del q},
$$
where total inertial $M(q) = \sum_i J_i^\T M_i J_i$ is simple to
compute, but the
Coriolis terms are more complicated.
}
%
but the Coriolis terms
$c_i$ are less obvious. Recursive Newton-Euler can be tuned to be
numerically faster, but does not provide this nice general and
invertible form.

In standard notation, general robot dynamics are written in the form

$$u = M(q)~ \ddot q + F(q, \dot q)$$

where, for multi-rigid-body systems, we derived $M(q) = \sum_i J_i^\T M_i
J_i$ and $F(q,\dot q) = \sum_i J_i^\T (c_i + g_i)$.

Keep in mind that only when $M(q)$ is invertible we have a one-to-one
relation between our controls and the system acceleration, and we have
the guarantee that our system accelerates with $\ddot q = M(q)^{-1} u -
F(q,\dot q)$ as desired. $M(q)$ is not invertible if $J_i$ do not have
full rank, e.g., if some body $i$ is not articulated at all and $J_i$
is zero. In this case the equation $u = J_i^\T \[M_i J_i \ddot q +
c_i\]$ says that we feel none of the accelerations of $i$ in our
joints -- and conversely cannot induce any accelerations of
$i$. That's the case when the robot is not in contact with body $i$.


%% That wrench has to come from connections to other bodies --
%% transmitted through the joints. We have the equation
%% \begin{align*}
%% F_i
%% &= F^\text{ext}_i + F^\text{joint}_i
%% - \sum_{j=\text{child(i)}} F^\text{joint}_j ~,
%% \end{align*}
%% where $F^\text{ext}_i$ are potential external forces (e.g.\ gravity, contacts) 
%% directly applying to $i$, and $F^\text{joint}_i$ is the force that the
%% parent $\pa(i)$ applies to $i$ through its joint. This equation is
%% recursive in all $F^\text{joint}_i$, and can easily be resolved
%% starting from leaf bodies. Once we have all joint wrenches, projecting
%% them to the joints' axes gives the torques $u_i = h_i^\T
%% F^\text{joint}_i$ (where $h_i\in\RRR^6$ is the ``joint axis'' in world
%% coordinates, generalized to include prismatic and/or rotational joints to
%% read out forces or torques depending on type).

%% Rather that implementing this recursion directly, we can also setup a
%% sparse linear equation system, that provides more options to be used:
%% We have $\ddot q, u\in\RRR^n$, matrices $F, F^e, F^j\in\RRR^{m\times
%%   6}$, and the equation systems
%% \begin{align}
%%  F = M J \ddot q + c\comma F = F^e + T F^j \comma u=H F^j ~,
%% \end{align}
%% where $M\in\RRR^{m \times 6 \times 6}$ contains $m_i \Id$ and $I_i$,
%% $c$ contains terms $w_i\times I_i w_i$, $F^e$ is defined externally,
%% and $T\in\RRR^{6m\times 6m}$ is very sparse, with $+1$ on the
%% diagonal, and some $-1$ in off-diagonals corresponding to child
%% relations in the tree. Putting this together we get
%% \begin{align}
%% u=H T^{-1}\[ M J \ddot q + c - F^e\] ~,
%% \end{align}
%% and we explicitly established the linear relation between $u$ and
%% $\ddot q$. Since matrices sparse and $T$ tree structured, inversion
%% and all multiplication operations are $O(m)$. Note, however, that all
%% quantities above are in world coordinates! This includes all inertia
%% matricies $I_i \get R_i I_i {R_i}^{-1}$, and joint axes $h_i$,
%% and $F^e$ includes implied torques relative to the world origin.


\subsection{Standard Usage: Waypoint + Reference Motion + Controller}


With the above equations we can accelerate the system in any way we
like -- at least those dofs that are currently articulable. In this
view, the rest is planning: We need to decide how we want to
accelerate the system right now in order to reach some future goal.

There are a myriad of opinions on how robotic control systems and middleware
should be structured. Here is just one version, which I consider a baseline.

Consider that we want to have the robot fulfill a kinematic constraint
$$\phi(q_{t=T}) = y^*$$ at time $t=T$, where $\phi$ is a $d$-dimensional
constraint function that typically depends on some poses $x_i$ of some bodies,
and $y^*\in\RRR^d$ is called a setpoint. A standard robotic system
could address this problem as follows:
\begin{enumerate}
  \item First compute a final robot pose $q_T$ that fulfills constraint 
    $\phi(q_{t=T}) = y^*$ -- that problem is called \textbf{inverse
    kinematics} and discussed below.
  \item Next compute a \emph{reference} motion from current robot pose
    $q_0$ to $q_T$ -- that problem can be addressed with \textbf{path
    finding}, \textbf{trajectory optimization}, or basic interpolation
    with a motion profile.
  \item Finally, determine a control policy $\pi: (x,t) \mapsto u$
    that reactively computes motor commands $u$ to follow the
    reference motion -- that problem can be addressed using \textbf{PD
    control}, inverse dynamics as derived above, \textbf{Riccati},
    or \textbf{model-predictive control (MPC)}.
\end{enumerate}
You could think of these as three different time scales: First rough
future waypoint(s)/goal(s), then continuous motion to next waypoint,
then short-term controls. Continuous replanning/re-estimation can also make (1) and (2) reactive.

Of course, robotics systems do not have to be organized in that
way: Some approaches skip step (1) and let step (2) also solve for the
final configuration (e.g., including the optimization of $q_T$ into
the trajectory optimization problem); or one may skip steps (1) and
(2) and let step (3) handle the full problem (e.g., including the goal
constraint in the MPC formulation, or a basic task-space PD controller
(``operational space control''); which typically looses the power of
path finding and optimization).


\subsection{Inverse Kinematics}

Inverse kinematics (IK) simply means computing $q$ to fulfill $\phi(q) =
y^*$. A proper approach is to formulate this as an NLP (non-linear
mathematical program)

\begin{align}
&\min_{q\in\RRR^n} \norm{q-q_0}^2 \st \phi(q) = y^* \label{eqIK}\\
\text{or}\quad&\min_{q\in\RRR^n} \norm{q-q_0}^2 + \m \norm{\phi(q) -
y^*}^2 \quad\text{for large $\m$} 
\end{align}

and use an efficient NLP solver (e.g.\ Augmented Lagrangian, or SQP,
exploiting potential sparseness of $\Del q \phi$). However, typical
textbooks at length discuss computing IK more low-level. For instance,
when approximating $\phi(q) \approx \phi(q_0) + J (q-q_0)$ as linear
with Jacobian $J$, the analytical solution to the above can be written
as (allowing for $\mu\to\infty$):

\begin{align*}
q^*
&= q_0 +  J^\T (J J^\T + \textstyle\frac{1}{\mu} \Id)^{-1} (y^*-\phi(q_0)) ~.
\end{align*}

In the context of optimization, this is the first Newton step when
initializing optimization at $q_0$. Students sometimes interpret this
equation as a tool to directly generate robot motion: They let the
robot literally execute these Newton steps (scaled by a small factor
$\a$), and then the robot starts moving like the decision variable in
a non-linear optimization problem with small-scaled Newton steps. This
is not proper! Proper IK should really first compute the solution
$q_T$ to
\eqref{eqIK}, and then think about how the robot can actually move
to $q_T$ (e.g.\ using proper optimal control, or reactive spline interpolation, or
a basic but nice motion profile).




\end{document}
